---
title: "Transportation Analysis | Passenger Commercial Flights"
author: "Alberto González Paje"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Unfortunately there is few places where to download and work with flights data. The only place I know is openflights.org, where you can doownload their example dataset and work with it. There is no free alternative as far as I know.


# Get data

There is three files to work with. Airlines, airports and routes.

```{r }
# Set wd
setwd("/home/albertogonzalez/Desktop/TDA/chapters/flights")

#Input data
airlines = read.csv("airlines.dat.txt",header = FALSE)

colnames(airlines) = c("id_air","name_air","alias_air","iata_air","icao_air","callsign_air","country_air","active_air")

head(airlines)


airports = read.csv("airports.dat.txt",header = FALSE)

colnames(airports) = c("id_port","name_port","city_port","country_port","iata_port","icao_port","lat_port","lon_port","alt_port","time_port","dst_port","timezone_port")

head(airports)


routes = read.csv("routes.dat.txt",header = FALSE)

colnames(routes) = c("name_air","id_air","name_air_orig","id_air_orig","name_air_dest","id_air_dest","codeshare","stops","equipment")

head(routes)


```

# Prepare data

Now I am going to merge the files so that we have a richer dataset.


```{r }
library(dplyr)
library(tidyr)
library(ggplot2)


airport_geo = airports[c("iata_port","lat_port","lon_port")]
head(airport_geo)

head(routes)

m1 = merge(routes,airport_geo,by.x ="name_air_orig",by.y = "iata_port")


colnames(m1) = c("name_air_orig", "name_air", "id_air", "id_air_orig", "name_air_dest","id_air_dest", "codeshare", "stops", "equipment", "lat_port_orig","lon_port_orig")

head(m1)

m2 = merge(m1,airport_geo,by.x ="name_air_dest",by.y = "iata_port")


colnames(m2) = c("name_air_dest","name_air_orig", "name_air", "id_air", "id_air_orig","id_air_dest", "codeshare", "stops", "equipment", "lat_port_orig","lon_port_orig","lat_port_dest","lon_port_dest")

head(m2)

```

# Routes flows

One of the first visual analysis should be in terms of routes. Let's plot all possible routes.Then, we can filter routes from a single airline and plot them.

```{r}
library(maps)
library(geosphere)


airlines_names = as.data.frame(table(m2$name_air)) %>%
  arrange(desc(Freq))

head(airlines_names,20)


#plot routes
head(routes)
head(airports)
head(m2)

orig = as.matrix(m2[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m2[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)

lines(i, col="black", lwd=0.1)

# filter by airline
head(airlines_names,20)

input_airline = "LH"

m3 = m2 %>%
  filter(name_air==input_airline)

orig = as.matrix(m3[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m3[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)

lines(i, col="red", lwd=0.1)


```


# Networks

We can consider airports as nodes and routes as relations amongst those nodes. Threfore we can work with those data in terms of a network. 
We calculate centrality metrics such as degree, betweenness and closeness and plot the top nodes according to the calculated metrics.

```{r}
library(igraph)
#library(Rgraphviz)

head(m2)

input_pair = m2[c("name_air_orig","name_air_dest")]
input_pair$count = 1
head(input_pair)


ne = graph.data.frame(input_pair)


# DEGREE (celebrities)
de = degree(ne)

summary(de)

top_in = as.data.frame(de)
str(top_in)
airport = row.names(top_in)
top_in = cbind(airport,top_in)
head(top_in)

top_in_ord = top_in %>%
  arrange(desc(de))

head(top_in_ord,20)

top_in_20 = top_in_ord[1]
top_in_20 = head(top_in_20,3)
top_in_20 = top_in_20$airport

head(m2)

m4 = m2 %>%
  filter(name_air_orig %in% top_in_20 | name_air_dest %in% top_in_20)

orig = as.matrix(m4[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m4[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)
lines(i, col="red", lwd=0.1)


# plot degree top 20 airlines routes

foo = function(i){
  fi = m2 %>% filter(name_air_orig %in% i | name_air_dest %in% i)
orig = as.matrix(fi[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(fi[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

ii = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)
lines(ii, col="red", lwd=0.1)
i
}

#head(top_in_ord,20)

foo("ATL")

inp = head(top_in_ord,20) 
input_list = inp$airport


foo_all = lapply(input_list,foo)
foo_all



# CLOSENESS (gossip - ability to receive and send information)
network_object = graph.data.frame(input_pair)
summary(network_object)
class(network_object)

clo = closeness(network_object)
summary(clo)
c = clo[1:20]

top_clo = as.data.frame(c)

airport = row.names(top_clo)
top_clo = cbind(airport,top_clo)
head(top_clo)

top_clo_ord = top_clo %>%
  arrange(desc(c))

top_clo_20 = top_clo_ord[1]
top_clo_20 = head(top_clo_20,10)
top_clo_20 = top_clo_20$airport

head(m2)

m5 = m2 %>%
  filter(name_air_orig %in% top_clo_20 | name_air_dest %in% top_clo_20)

orig = as.matrix(m5[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m5[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)
lines(i, col="red", lwd=0.1)


##plot 10 top airports by closeness
input_list_clo = top_clo_20
foo_all_clo = lapply(input_list_clo,foo)
foo_all_clo




# BETWEENEESS (bottlenecks and comunity bridges)
bet = betweenness(network_object)
b = bet[1:20]
length(bet)

b


top_bet = as.data.frame(b)

airport = row.names(top_bet)
top_bet = cbind(airport,top_bet)
head(top_bet)

top_bet_ord = top_bet %>%
  arrange(desc(b))

top_bet_20 = top_bet_ord[1]
top_bet_20 = head(top_bet_20,10)
top_bet_20 = top_bet_20$airport

head(top_bet_ord)

head(m2)

m6 = m2 %>%
  filter(name_air_orig %in% top_bet_20 | name_air_dest %in% top_bet_20)

orig = as.matrix(m6[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m6[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)
lines(i, col="red", lwd=0.1)

##plot 10 top airports by betweeneess
input_list_bet = top_bet_20
foo_all_bet = lapply(input_list_bet,foo)
foo_all_bet



```


# Clusters

Finally we can apply cluster analysis. We will first calculate the number of optimal clusters and then apply that input to a k-means algo.

```{r}

ne
head(m2)


# cluster airports by position
air_po = m2[c("name_air_orig","lat_port_orig","lon_port_orig")]
head(air_po)
dim(air_po)

air_po_u = unique(air_po)
dim(air_po_u)
head(air_po_u)

plot(air_po_u$lon_port_orig,air_po_u$lat_port_orig)

mydata = air_po_u[2:3]

wss <- (nrow(mydata)-1)*sum(apply(mydata,2,var))
  for (i in 2:15) wss[i] <- sum(kmeans(mydata,
                                       centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares")


library(fpc)
pamk.best <- pamk(mydata)

summary(pamk.best)

cat("number of clusters estimated by optimum average silhouette width:", pamk.best$nc, "\n")


clus = kmeans(mydata,2)
clus_id = as.data.frame(clus$cluster)
dim(clus_id)

clus_df = cbind(air_po_u,clus_id)

colnames(clus_df) = c("air","lat","lon","clu")
head(clus_df,20)


#plot cluster routes
clu_1 = clus_df %>%
  filter(clu==1)

head(clu_1)

clu_1_names = clu_1$air
head(clu_1_names)

head(m2)

m7 = m2 %>%
  filter(name_air_orig %in% clu_1_names)

head(m7)

orig = as.matrix(m7[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m7[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)
lines(i, col="red", lwd=0.1)


clu_2 = clus_df %>%
  filter(clu==2)

head(clu_2)

clu_2_names = clu_2$air
head(clu_2_names)

head(m2)

m8 = m2 %>%
  filter(name_air_orig %in% clu_2_names)

head(m8)

orig = as.matrix(m8[c("lon_port_orig","lat_port_orig")])
dest = as.matrix(m8[c("lon_port_dest","lat_port_dest")])

inter = gcIntermediate(orig,dest)

i = do.call("rbind", inter)

map("world", col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05)
lines(i, col="red", lwd=0.1)


```
























